### 2022-03-26 ehoh2/ehoh2.tex
more

\begin{abstract}%
    The automatic discharge of tedious subgoals is high on the wishlist of many
  users of proof assistants. Some proof assistants discharge such goals
  by translating them to first-order logic and invoking an efficient prover on
  them, but much is lost in translation. As an alternative,
  we propose to extend first-order provers with native support for
  higher-order features. Building on our extension of E to $\lambda$-free
  higher-order logic, we now extend E to full higher-order logic.
  The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
  proof assistant, and the second best on TPTP benchmarks.
    %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
\end{abstract}

@@

We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:ehoh2:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on
proof assistant benchmarks; on TPTP benchmarks it ended up second only 
to the cooperative version of Zipperposition, which employs Ehoh as a
backend. An arguably fairer comparison without the backend puts \ehohii{} in the
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm} which is set if
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

\looseness=-1
More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
the cell will be of the form
$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
where $t$ is not a
$\lambda$-abstraction, and $l$ is the number of $\internallam$ symbols above $t$. Then \ehohii{} breaks down the body~$t$ into a maximal
decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
If $n = 0$, the cell is not $\eta$-reducible.
Otherwise, $u$ is traversed to determine the
minimal index~$\dbvar{j}$ of a loose De Bruijn index,
taking $\dbvar{j} = \infty$ if no such index exists.
%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
%bottom-up as innermost can suggest): Check "innermost".
%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
%%% binder remains (the one to which the 2's point). --JB
We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
$t$ by the variant of
$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
obtained by shifting the loose De Bruijn indices right by~$k$.

@@

Computing bindings is the only point in the procedure where the search tree
branches and different possibilities are explored. Thus, when \ehohii{} follows
the branch indicated by the binding module, it records the state to which it
needs to return, should the followed branch be backtracked. It saves the values
of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
and the value of the \vn{branch\_iter} that points past the followed branch. The
values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
binding was created, intermediate values that \cn{NextBinding} uses to remember
how far through bindings it is, and \cn{BindEnd}, which indicates that all
bindings are exhausted.

@@

  %TODO: @JASMIN: How do we cite the STTT article?
  The third difference is more subtle. One of the main features of Ehoh is
  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): A method which, given a demodulator $s \eq t$, allows
  to replace both applied and unapplied occurrences of $s$ by $t$ 
  by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
  optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
  terms can change drastically, making it much harder to
  track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
  To counterbalance this removal, we introduced
  the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
  in \ehohii{}
  and enabled positive and negative functional extensionality (\infname{PE} and
  \infname{NE}) by default:

@@

  In this section, we try to answer two questions about \ehohii{}: \emph{How does
  \ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
  \ehohii{} introduce any overhead compared with Ehoh?} To answer
  these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
  benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
  The experiments were carried out on StarExec Miami
  \cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
  2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
  wallclock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
  default time limit:\ 30~s wallclock and CPU.

@@

  \ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
  excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
  If Zipperposition's Ehoh backend is disabled,
  \ehohii{} outperforms Zipperposition by a wide margin.
  This comparison is arguably fairer; after all,
  \ehohii{} does not use an older version of Zipperposition
  as a backend.
  %
  These results suggest that \ehohii{} already implements most of the necessary features
  for a high-performance higher-order prover
  but could benefit from the kind of fine-tuning that
  Zipperposition underwent in the last two years.

@@

                              & TPTP       & SH        \\
        \midrule
        cvc5                  & 1931       & 2577      \\
        Ehoh                  & 2105       & 2611      \\
        \ehohii{}             & 2523       & {\bf 2766}\\
        Leo-III               & 2282       & 1601      \\
        Satallax              & 2320       & 1719      \\
        Vampire               & 2203       & 2240      \\
        Zipperposition-coop   & {\bf 2583} & 2754      \\
        Zipperposition-uncoop & 2483       & 2181      \\
        \bottomrule
      \end{tabular}}
      \caption{Comparison of higher-order provers}
      \label{fig:all-provers}
    \end{minipage}\hfill\hfill
    \begin{minipage}[ht!]{.30\linewidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
                        & TPTP     \\
        \midrule
        Ehoh -- FO      & 535      \\
        Ehoh -- HO      & 538      \\
        \ehohii{} -- FO & {\bf 539}\\
        \ehohii{} -- HO & {\bf 539}\\
        \bottomrule
      \end{tabular}}
      \caption{Evaluation of \ehohii{}'s overhead}
      \label{fig:ehoh}
    \end{minipage}\hfill\hfill
  \end{figure}

@@

  In 2019, the reviewers of our paper introducing Ehoh \cite{vbss-19-ehoh1} were
  skeptical that extending it with support for full higher-order logic would be
  feasible. One of them wrote:
  %
  \begin{quote}
  A potential criticism could be that this step from E to Ehoh is just extending
  FOL by those aspects of HOL that are easily in reach with rather straightforward
  extensions (none of the extensions is indeed very complicated), and that the
  difficult challenges of fully supporting HOL have yet to be confronted.
  \end{quote}
  %
  We ended up addressing the theoretical ``difficult challenges'' in other work
  with colleagues. In this chapter, we faced the practical challenges pertaining to the extension
  of Ehoh's data structures and algorithms to support full higher-order logic and
  demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
  that this extension makes \ehohii{} the best higher-order prover on
  benchmarks coming from interactive theorem proving practice, which has been our goal since 2016. \ehohii{}'s
  performance lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
  does not assume a clausal structure and can perform subtle formula-level
  inferences. In the future, we plan to implement the same features in \ehohii{}.
  We have also only started tuning \ehohii{}'s heuristics on higher-order
  problems.

>>>>

\begin{abstract}%
    The automatic discharge of tedious subgoals is high on the wishlist of many
  users of proof assistants. Some proof assistants discharge such goals
  by translating them to first-order logic and invoking an efficient prover on
  them, but much is lost in translation. As an alternative,
  we propose to extend first-order provers with native support for
  higher-order features. Building on our extension of E to $\lambda$-free
  higher-order logic, we now extend E to full higher-order logic.
  The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
  proof assistant, and the \NumberOK{second strongest} on TPTP benchmarks.
    %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
\end{abstract}

@@

We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:ehoh2:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on the
proof assistant benchmarks; on the TPTP benchmarks it ended up second only 
to the cooperative version of Zipperposition, which employs Ehoh as a
backend. An arguably fairer comparison without the backend puts \ehohii{} in the
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm}, which is set if
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

\looseness=-1
More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
the cell will be of the form
$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
where $t$ is not a
$\lambda$-abstraction, and $l$ is the number of $\internallam$ symbols above $t$. Then \ehohii{} breaks down the body~$t$ into a maximal
decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
If $n = 0$, the cell is not $\eta$-reducible.
Otherwise, $u$ is traversed to determine the
minimal index~$\dbvar{j}$ of a loose De Bruijn index,
taking $\dbvar{j} = \infty$ if no such index exists.
%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
%bottom-up as innermost can suggest): Check "innermost".
%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
%%% binder remains (the one to which the 2's point). --JB
We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
$t$ by the variant of
$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
obtained by shifting the loose De Bruijn indices down by~$k$.

@@

Computing bindings is the only point in the procedure where the search tree
branches and different possibilities are explored. Thus, when \ehohii{} follows
the branch indicated by the binding module, it records the state to which it
needs to return, should the followed branch be backtracked. The state consists of
the values
of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
and the value of \vn{branch\_iter} that points past the followed branch. The
values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
binding was created, intermediate values that \cn{NextBinding} uses to remember
how far through bindings it is, and \cn{BindEnd}, which indicates that all
bindings are exhausted.

@@

  %TODO: @JASMIN: How do we cite the STTT article?
  The third difference is more subtle. One of the main features of Ehoh is
  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): a method that, given a demodulator $s \eq t$, allows
  to replace both applied and unapplied occurrences of $s$ by $t$ 
  by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
  optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
  terms can change drastically, making it much harder to
  track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
  To counterbalance this removal, we introduced
  the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
  in \ehohii{}
  and enabled positive and negative functional extensionality (\infname{PE} and
  \infname{NE}) by default:

@@

  In this section, we try to answer two questions about \ehohii{}: \emph{How does
  \ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
  \ehohii{} introduce any overhead compared with Ehoh?} To answer
  these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
  benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
  The experiments were carried out on StarExec Miami
  \cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
  2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
  wall-clock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
  default time limit:\ 30~s wall-clock and CPU.

@@

  \ehohii{} was mainly designed as an efficient backed to proof assistants. As such it 
  excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
  If Zipperposition's Ehoh backend is disabled,
  \ehohii{} outperforms Zipperposition by a wide margin.
  This comparison is arguably fairer; after all,
  \ehohii{} does not use an older version of Zipperposition
  as a backend.
  %
  These results suggest that \ehohii{} already implements most of the necessary features
  for a high-performance higher-order prover
  but could benefit from the kind of fine-tuning that
  Zipperposition underwent in the last two years.

@@

                              & TPTP       & SH        \\
        \midrule
        cvc5                  & 1931       & 2577      \\
        Ehoh                  & 2105       & 2611      \\
        \ehohii{}             & 2523       & {\bf 2766}\\
        Leo-III               & 2282       & 1601      \\
        Satallax              & 2320       & 1719      \\
        Vampire               & 2203       & 2240      \\
        Zipperposition-coop   & {\bf 2583} & 2754      \\
        Zipperposition-uncoop & 2483       & 2181      \\
        \bottomrule
      \end{tabular}}
      \caption{Comparison of higher-order provers}
      \label{fig:all-provers}
    \end{minipage}\hfill\hfill
    \begin{minipage}[ht!]{.30\linewidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
                        & TPTP     \\
        \midrule
        Ehoh FO      & 535      \\
        Ehoh HO      & 538      \\
        \ehohii{} FO & {\bf 539}\\
        \ehohii{} HO & {\bf 539}\\
        \bottomrule
      \end{tabular}}
      \caption{Evaluation of \ehohii{}'s overhead}
      \label{fig:ehoh}
    \end{minipage}\hfill\hfill
  \end{figure}

@@

  In 2019, the reviewers of our paper introducing Ehoh \cite{vbss-19-ehoh1} were
  skeptical that extending it with support for full higher-order logic would be
  feasible. One of them wrote:
  %
  \begin{quote}
  A potential criticism could be that this step from E to Ehoh is just extending
  FOL by those aspects of HOL that are easily in reach with rather straightforward
  extensions (none of the extensions is indeed very complicated), and that the
  difficult challenges of fully supporting HOL have yet to be confronted.
  \end{quote}
  %
  We ended up addressing the theoretical ``difficult challenges'' in other work
  with colleagues. In this chapter, we faced the practical challenges pertaining to the extension
  of Ehoh's data structures and algorithms to support full higher-order logic and
  demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
  that this extension makes \ehohii{} the best higher-order prover on
  benchmarks coming from interactive theorem proving practice, which was our goal. \ehohii{}
  lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
  does not assume a clausal structure and can perform subtle formula-level
  inferences. In the future, we plan to implement the same features in \ehohii{}.
  We have also only started tuning \ehohii{}'s heuristics on higher-order
  problems.
